---
title: "Assignment 2: Coding Basics"
author: "Joanna Huertas"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(error = TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. seq(1, 10)
fours_sequence <- seq(1, 100, 4) # from, to, by
fours_sequence

#2. 
mean<-mean(fours_sequence) #finding the mean, assigning a name 
median<-median(fours_sequence) #finding the median, assigning a name 
mean
median

#3.
mean>median #asking R for True or False

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#creating a series of vectors
studentnames <- c('Anna','Tom','Lucas','Lucy') #character
tests <- 90:93 #sequence
# Which scores are >= 50?
results <- tests >= 50 #logical
results

students_data <- data.frame(studentnames, tests, results)    # Apply data.frame function
students_data

colnames(students_data) <- c('Student Names','Test Scores','Pass')
students_data
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:In a data frame the columns contain different types of data, but in a matrix all the elements are the same type of data. A matrix in R is like a mathematical matrix, containing all the same type of thing (usually numbers). 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}

#trying with 'if' and 'else'
pass <- function(x) {
  if(x >= 50) {
    'TRUE'
  } 
  else {
    'FALSE'
  }
}


pass_or_not <- pass(tests); pass_or_not
pass_or_not 


#trying with 'ifelse'
pass1 <- function(x){
  ifelse(x >= 50, 'TRUE', 'FALSE') #log_exp, if TRUE, if FALSE
  # 
}

pass_or_not1 <- pass1(tests); pass_or_not1
pass_or_not1 
#this works

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:`ifelse` worked because the 'if' function is designed to work with things that are length 1, like a single name, which is why we got an error using it with 4 elements; while if we want to work with vectors that are length > 1, like in this case, we use 'ifelse'.
