---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Joanna Huertas"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW
```{r setup, include=FALSE}

```

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(here)
library(tidyverse)
library(agricolae)
library(corrplot)
library(lubridate)
here()

NTL_LTER <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors = TRUE)

# convert the "Date" column to a date object
NTL_LTER$sampledate <- as.Date(NTL_LTER$sampledat, format = "%m/%d/%y")

#2
# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer: Yes, as we can see later in the graphic, there is a negative correlation between temperature and depth.
H0: depth across all lakes has no correlation with the temperature 
Ha: depth across all lakes has correlation with the temperature 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

NTL_LTER.wrangle <- 
  NTL_LTER %>%
  filter(month(sampledate)==7)%>%
  select(lakename, year4, daynum, depth, temperature_C)%>%
  drop_na()


#5
scatterplot <- 
  ggplot(NTL_LTER.wrangle, aes(x=depth, y=temperature_C, color = lakename))+
  geom_point()+
  geom_smooth(method="lm", se = FALSE, color = "blue") +
  #xlim(0, 125) +
  ylim(0, 35)+
  labs(title= "Temperature vs Depth", x= "Depth (m)", y= "Temperature (°C)", color="Lakes")+
   theme(legend.text = element_text(size = 7), legend.position = "top", 
         legend.justification = "right", legend.key.width = unit(10, "pt"),
         legend.key.height = unit(1, "cm"))

print(scatterplot)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: In this case, the scatterplot suggests that temperature is a function of depth, with temperature decreasing as depth increases, indicating a strong negative relationship. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

temp.depth.regression <- lm(data = NTL_LTER.wrangle, depth ~ temperature_C)
summary(temp.depth.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:The linear regression analysis indicates that changes in depth explain 73.87% of the variability in temperature (notice that R-squared is close to 1). This finding is based on 1 degree of freedom and 9726 degrees of freedom for the residuals, with a residual standard error of 1.694. The F-statistic for the regression is 2.75e+04 with a p-value of < 2.2e-16 (p<alpha), indicating that the regression is statistically significant.
Change in temperature = -0.379578 x 1 = -0.379578°C
Therefore, for every 1 meter increase in depth, temperature is predicted to decrease by approximately 0.38°C, according to the linear regression analysis.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

cor.test(NTL_LTER.wrangle$depth, NTL_LTER.wrangle$temperature_C)

TPAIC.1 <- lm(data = NTL_LTER.wrangle, temperature_C ~ year4 + daynum + depth)

#Choose a model by AIC in a Stepwise Algorithm
step(TPAIC.1)

#10
TPmodel <- lm(data = NTL_LTER.wrangle, temperature_C ~ year4 + daynum + 
              depth)
summary(TPmodel)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer:  The AIC method suggests using all three explanatory variables, "year4", "daynum", and "depth", to predict temperature in the multiple regression.
The model explains 100% - (141687 / 26066) = 81.54% of the observed variance in the data.The initial AIC value was 26065.53, and the model including all three explanatory variables had an AIC of 26066, which is only slightly higher. This suggests that the additional explanatory variables have not significantly improved the model's fit.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# Wrangle the data
NTL_LTER.temperatures <- NTL_LTER.wrangle %>%
  group_by(year4, daynum, lakename) %>%
  summarise(temperature_C = mean(temperature_C))

summary(NTL_LTER.temperatures)
#check all value of lakename
summary(NTL_LTER.temperatures$lakename)

# Format ANOVA as aov
NTL_LTER.temperatures.anova <- aov(data = NTL_LTER.temperatures, temperature_C ~ lakename)
summary(NTL_LTER.temperatures.anova)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes. The P value is less than 0.05, then we reject the null hypothesis (the mean is the same across all different lakes), this the means are not all the same.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

scatterplot2 <- 
ggplot(NTL_LTER.wrangle, aes(x=depth, y=temperature_C, color = lakename))+
  geom_point(alpha = 0.5)+
geom_smooth(method="lm", color="blue", se=FALSE)+
  #xlim(0, 125) +
  ylim(0, 35)+
  labs(fill="", title= "Temperature vs Depth", x= "Depth (m)", y= "Temperature (°C)", color="Lakes")+
   theme(legend.text = element_text(size = 7), legend.position = "top", 
         legend.justification = "right", legend.key.width = unit(10, "pt"),
         legend.key.height = unit(1, "cm"))

print(scatterplot2)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
# Post-hoc test
# TukeyHSD() computes Tukey Honest Significant Differences
TukeyHSD(NTL_LTER.temperatures.anova)

# Extract groupings for pairwise relationships
NTL_LTER.temperatures.groups <- HSD.test(NTL_LTER.temperatures.anova, "lakename", group = TRUE)
NTL_LTER.temperatures.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:Paul Lake has the same mean temperature as Peter Lake. Central Long Lake has a mean temperature that is statistically distinct from all the other lakes.

 
17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we are only interested in comparing the mean temperatures of Peter Lake and Paul Lake, we could use a two-sample t-test to determine whether their means are statistically different.

18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
# Wrangle the data
NTL_LTER.Crampton.Ward.Lakes <- NTL_LTER.wrangle %>%
  filter(lakename== "Crampton Lake" | lakename== "Ward Lake")%>%
  group_by(year4, daynum, lakename) %>%
  summarise(temperature_C = mean(temperature_C))

summary(NTL_LTER.Crampton.Ward.Lakes)
#check all value of lakename
summary(NTL_LTER.Crampton.Ward.Lakes$lakename)


#Format as a t-test
#EPAair$Ozone will be our continuous dependent variable
#EPAair$Year will be our categorical variable with two levels (2018 and 2019)
Crampton.Ward.twosample <- t.test(NTL_LTER.Crampton.Ward.Lakes$temperature_C ~
                                    NTL_LTER.Crampton.Ward.Lakes$lakename)
Crampton.Ward.twosample

# Format as a GLM
Crampton.Ward.twosample2 <- lm(NTL_LTER.Crampton.Ward.Lakes$temperature_C ~
                                 NTL_LTER.Crampton.Ward.Lakes$lakename)
summary(Crampton.Ward.twosample2)
```

>Answer: Our p-value is higher than 0.05 so we don't reject the null hypothesis, therefore, the two means (Crampton Lake and Ward Lake temperatures) are the same.My answer in part 16 says that Crampton Lake and Ward Lake have similar means but not the same, because Crampton Lake has the group b and Ward Lake has the group bc, so they only share b.
