---
title: "Assignment 8: Time Series Analysis"
author: "Joanna Huertas"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_TimeSeries.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

```{r}
#1
getwd()

library(tidyverse)
library(lubridate)
library(trend)
library(zoo)
library(Kendall)
library(tseries)
library(dplyr)
library(forecast)


# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```


2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r, message = FALSE}
#2
#Import Ozone_TimeSeries
folder_contents <- list.files("./Data/Raw/Ozone_TimeSeries")
folder_contents

EPA2010 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2010_raw.csv",
                         stringsAsFactors = TRUE)
EPA2011 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2011_raw.csv",
                         stringsAsFactors = TRUE)
EPA2012 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2012_raw.csv",
                         stringsAsFactors = TRUE)
EPA2013 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2013_raw.csv",
                         stringsAsFactors = TRUE)
EPA2014 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2014_raw.csv",
                         stringsAsFactors = TRUE)
EPA2015 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2015_raw.csv",
                         stringsAsFactors = TRUE)
EPA2016 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2016_raw.csv",
                         stringsAsFactors = TRUE)
EPA2017 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2017_raw.csv",
                         stringsAsFactors = TRUE)
EPA2018 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2018_raw.csv",
                         stringsAsFactors = TRUE)
EPA2019 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2019_raw.csv",
                         stringsAsFactors = TRUE)


GaringerOzone <- rbind(EPA2010, EPA2011, EPA2012, EPA2013, EPA2014, EPA2015, EPA2016,
                       EPA2017, EPA2018, EPA2019)


```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}
#3

GaringerOzone$Date <- as.Date(GaringerOzone$Date, format = "%m/%d/%Y")

#4

GaringerOzone.select <- 
  GaringerOzone %>%
  select(Date, Daily.Max.8.hour.Ozone.Concentration, DAILY_AQI_VALUE)

#5

#GaringerOzone.select.1 <- replace(GaringerOzone.select, is.na(GaringerOzone.select$Daily.Max.8.hour.Ozone.Concentration), NA)

Days <- as.data.frame(seq(as.Date("2010-01-01"), as.Date("2019-12-31"), by = "day"))
colnames(Days)[1] <- "Date"

#6

GaringerOzone <- left_join(Days, GaringerOzone.select, by = "Date")

```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r}
#7

ggplot(GaringerOzone, aes(x = Date, y = Daily.Max.8.hour.Ozone.Concentration
                          )) +
  geom_line() +
  labs(x = "Time", y = expression("Ozone (ppm)"))+
  geom_smooth(method = 'lm')


```

>Answer: Yes, the trend line is slightly downward over time.

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r}
#8
summary(GaringerOzone$Daily.Max.8.hour.Ozone.Concentration)
summary(GaringerOzone$DAILY_AQI_VALUE)
#We have 63 NA's

# replace NAs
GaringerOzone.clean <- 
  GaringerOzone %>% 
  mutate(Ozone.Concentration.clean = zoo::na.approx(Daily.Max.8.hour.Ozone.Concentration))

summary(GaringerOzone.clean$Ozone.Concentration.clean)
#NA is gone

ggplot(GaringerOzone.clean) +
  geom_line(aes(x = Date, y = Ozone.Concentration.clean), color = "red") +
  geom_line(aes(x = Date, y = Daily.Max.8.hour.Ozone.Concentration), color = "black") +
  ylab("Ozone concentration") 
```

> Answer: yes, 
linear interpolation is a suitable method for estimating missing daily data for ozone concentration because it assumes a linear relationship between data points, is simple to implement, and is appropriate for continuous and smooth data. Piecewise constant interpolation is not appropriate for this type of data because ozone concentration is unlikely to change abruptly. Spline interpolation may be unnecessary for relatively simple data sets and requires more computational power, it uses cuadratic function to interpolate.


9. Create a new data frame called `GaringerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r}
#9
GaringerOzone.monthly <- 
  GaringerOzone.clean %>%
  mutate(year = year(Date), 
         month = month(Date, label = TRUE)) %>%
  group_by(year, month) %>%
  summarize(mean_ozone = mean(Ozone.Concentration.clean, na.rm = TRUE))

GaringerOzone.monthly$Date <- as.Date(paste(GaringerOzone.monthly$year, 
                                            as.numeric(GaringerOzone.monthly$month), 
                                            "01", 
                                            sep = "-"))
GaringerOzone.monthly

```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r}
#10
f_month <- month(first(GaringerOzone.clean$Date))
f_year <- year(first(GaringerOzone.clean$Date))
f_month
f_year
GaringerOzone.daily.ts <- ts(GaringerOzone.clean$Ozone.Concentration.clean,
                   start=c(2010,1),
                   frequency=365) 

summary(GaringerOzone.daily.ts)
GaringerOzone.monthly.ts <- ts(GaringerOzone.monthly$mean_ozone,
                   start=c(2010,1),
                   frequency=12) 
summary(GaringerOzone.monthly.ts)
head(GaringerOzone.monthly.ts)
GaringerOzone.monthly.ts
```


11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r}
#11
#decompose
GaringerOzone.daily_decomp <- stl(GaringerOzone.daily.ts,s.window = "periodic") #frequency 12
plot(GaringerOzone.daily_decomp)

GaringerOzone.monthly_decomp <- stl(GaringerOzone.monthly.ts,s.window = "periodic") #frequency 12, periodic
plot(GaringerOzone.monthly_decomp)

```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r}
#12

# Run SMK test
GaringerOzone.monthly_trend <- Kendall::SeasonalMannKendall(GaringerOzone.monthly.ts)

# Inspect results
GaringerOzone.monthly_trend
summary(GaringerOzone.monthly_trend)


GaringerOzone.monthly_trend1 <- trend::smk.test(GaringerOzone.monthly.ts)
# Inspect results
GaringerOzone.monthly_trend1
summary(GaringerOzone.monthly_trend1)

```

> Answer: Because Mann-Kendall is appropiate for seasonal data, and the monthly Ozone series suggests seasonal data

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r}
#13

#Visualization
GaringerOzone.monthly_plot <-
ggplot(GaringerOzone.monthly, aes(x = Date, y = mean_ozone)) +
  geom_point() +
  geom_line() +
  scale_x_date(date_labels = "%Y", date_breaks  ="1 year")+
  ylab("Ozone concentration in ppm") +
  geom_smooth( method = lm )
print(GaringerOzone.monthly_plot)


```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: in the first method of SeasonalMannKendall, we notice the p-value is 
less than 0.05, so we are going to reject the null hypothesis that
the data is stationary. In the second method, we notice S is high in most cases,
which indicates stronger tendency of decrease, and in other cases the trend
is stationary but overall the trend is decrease.
The plot shows a negative slope.

15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r}
#15

# We can extract the components and turn them into data frames
GaringerOzone.monthly.ts_Components <- as.data.frame(GaringerOzone.monthly_decomp$time.series[,1:3])

GaringerOzone.monthly.ts_Components <- mutate(GaringerOzone.monthly.ts_Components,
        Observed = GaringerOzone.monthly$mean_ozone,     
        Date = GaringerOzone.monthly$Date)

GaringerOzone.monthly.ts_Components

# Subtract the seasonal component from the original time series
GaringerOzone.monthly_no_seasonality <- GaringerOzone.monthly.ts - 
  GaringerOzone.monthly_decomp$time.series[, "seasonal"]

#16
GaringerOzone.monthly_trend1 <- Kendall::MannKendall(GaringerOzone.monthly_no_seasonality)
# Inspect results
GaringerOzone.monthly_trend1
summary(GaringerOzone.monthly_trend1)


```

> Answer: In this case, the p-value obtained from the Seasonal Mann Kendall test on the complete ozone monthly series is 0.046, which suggests that there is a significant trend in the series, but this trend may be due to seasonality. On the other hand, the p-value obtained from the non-seasonal ozone monthly series is 0.0075402, which suggests that there is a significant trend in the series even after the seasonal component has been removed. Since 0.0075402 is less than 0.046, the first one is way more significant. Comparing both p-values, the trend detected by the Seasonal Mann Kendall test on the complete series may be partly explained by seasonality.
